A tool for designing digital filters for the Hankel and Fourier transforms in potential, diffusive, and wavefield modeling

### 2018-06-08 ManuscriptGeophysics2018.tex
Add number (GEO-2018-0069)
In review; pseudo-code

% \documentclass[manuscript]{geophysics}
\documentclass[paper,twocolumn,twoside]{geophysics}
% \documentclass[paper]{geophysics}
% \documentclass[paper, revised]{geophysics}
% \documentclass[manuscript,revised]{geophysics}

@@

\ms{GEO-2018-????}

\address{
\footnotemark[1]Instituto Mexicano del Petróleo,
Eje Central Lázaro Cárdenas Norte 152,
Col. San Bartolo Atepehuacan C.P. 07730,
Ciudad de México, México,
E-mail: \href{mailto:dieter@werthmuller.org}{Dieter@Werthmuller.org};
\footnotemark[2]Lamont-Doherty Earth Observatory, Columbia University,
305C Oceanography,
61 Route 9W, PO Box 1000,
Palisades NY 10964-8000 US,
E-mail: \href{mailto:KKey@ldeo.columbia.edu}{KKey@ldeo.columbia.edu};
\footnotemark[3]TU Delft,
Building 23,
Stevinweg 1 / PO-box 5048,
2628 CN Delft,
E-mail: \href{mailto:E.C.Slob@tudelft.nl}{E.C.Slob@tudelft.nl}.}

\author{%
Dieter Werthmüller\footnotemark[1], %
Kerry Key\footnotemark[2], and %
Evert C.\ Slob\footnotemark[3]%
}

\footer{}
\lefthead{Werthmüller et al.}
\righthead{Digital filter designing tool}

@@

By substituting $r = e^x$ and $l = e^{-y}$ we get
%
\begin{equation}
  e^x F(e^x) = \int^\infty_{-\infty} f(e^{-y})K(e^{x-y})e^{x-y}\mr{d}y\ .
  \label{eq:filtint}
\end{equation}
%
This can be re-written as a convolution integral and be approximated for an
$N$-point filter by
%
\begin{equation}
  F(r) \approx \sum^N_{n=1} \frac{f(b_n/r) h_n}{r}\ ,
  \label{eq:filtapprox}
\end{equation}
%
where $h$ is the digital linear filter, and the logarithmically spaced filter
abscissae is a function of the spacing $\Delta$ and the shift $\delta$,
%
\begin{equation}
  b_n = \exp\left\{\Delta(-N/2+n) + \delta\right\} \ .
  \label{eq:base}
\end{equation}
%
From equation~\ref{eq:filtapprox} it can be seen that the filter method
requires $N$ evaluations at each $r$. For example, to calculate the frequency
domain result for 100 offsets with a 201\,pt filter requires 20,100 evaluations
in the wavenumber domain. This is why the DLF often uses interpolation to
minimize the required evaluations, either for $F(r)$ in what is referred to as
\emph{lagged convolution DLF}, or for $f(l)$, which we here call \emph{splined
DLF}.

In the direct matrix inversion method for solving for the digital filter
coefficients, equation~\ref{eq:filtapprox} is cast as a linear system as
follows. $\Delta$ and $\delta$ are preassigned scalar values and $r$ is a range
of $M$ preassigned values $r_m$. The filter base coefficients $b_n$ are
computed using equation~\ref{eq:base} and an array of values for $l$ are
computed using $l_{mn} = b_n/r_m$. The linear system's left hand side (LHS)
matrix $\bf{A}$ has dimensions $M \times N$ with coefficients $A_{mn} =
f(l_{mn})/r_n$ and the right hand side (RHS) vector $\bf{v}$ has elements $v_n
= F(r_n)$. The resulting linear system
%
\begin{equation}
  \bf{A}\bf{h} = \bf{v}
  \label{eq:,linearSystem}
\end{equation}
%
can be solved by direct matrix inversion to obtain the vector of filter
coefficients $\bf{h}$. For a given filter length $N$, there are many subjective
choices that go into the practical implementation of this method, including the
choice of the transform pairs $F(r)$ and $f(l)$, as well as the value for $M$
and the particular range and spacing of the values $r_m$. Once values are
chosen for these variables, an optimal filter can be found by a grid search
over $\Delta$ and $\delta$ for values that produce a high quality filter. The
choice of metric for what constitutes a high quality filter is also subjective
and is further discussed below.

@@

The basic steps are as follows:
\begin{enumerate}
  \item Evaluate RHS of check-function $f_\mr{C}$:\newline
        $d_\mr{R} = f_\mr{C}.\mr{rhs}(r)$
  \item Loop over each value $\Delta_i$, $\delta_j$ (brute force):
    \begin{enumerate}
      \item Calculate filter base:\newline
        $b_n = \exp\left\{\Delta_i(-N/2+n) + \delta_j\right\}$
      \item Get required RHS and LHS evaluation points:\newline
        $r_\mr{I} = p(b, N)$\newline
        $l_\mr{I} = b/r_\mr{I}$
      \item Invert for filter coefficients:\newline
        $J_{ij} = \mr{solve}(f_\mr{I}.\mr{lhs}(l_\mr{I}),
        f_\mr{I}.\mr{rhs}(r_\mr{I}))$
    \item Calculate numerically RHS of check-function $f_\mr{C}$ with current
      filter $J_{ij}$:\newline
      $d_\mr{F} = f_\mr{C}.\mr{lhs}(b/r)\cdot J_{ij}\ /\ r$
      \item Store minimum recovered amplitude or maximum $r$ where relative
        error is less than the provided error:\newline
        $\chi_{ij} = g\left[
        \mr{argmin}\left(|(d_\mr{F}-d_\mr{R})/d_\mr{R}| >
        \mr{error}\right) -1\right]\ $, where $g$ is either $d_\mr{R}$
        or $1/r$.
    \end{enumerate}
  \item Return filter coefficients which yield minimum amplitude or maximum $r$
    (a local minimization can be run to polish the brute-force result):\newline
    return $J\left[\mr{argmin}(\chi)\right] $\newline
\end{enumerate}

@@

The presented, free and open-source code \texttt{fdesign} can be used to design
digital linear filters for the Hankel and Fourier transforms (and more
generally for any linear transform) using either analytical transform pairs or
1D subsurface models together with the EM-modeler \texttt{empymod}. The code is
available from GitHub in the add-on repository of \texttt{empymod}, called
\texttt{empyscripts}, from version 0.2.0 onwards.

@@

D.W. would like to thank the electromagnetic research group of Aleksandr
Mousatov at the IMP for fruitful discussions, and Jürg Hunziker for valuable
feedback which improved this manuscript considerably.

>>>>

% \documentclass[manuscript]{geophysics}
% \documentclass[paper,twocolumn,twoside]{geophysics}
% \documentclass[paper]{geophysics}
% \documentclass[paper, revised]{geophysics}
\documentclass[manuscript,revised]{geophysics}

@@

\ms{GEO-2018-0069.R2}

\address{
% \footnotemark[1]TU Delft,
% Building 23,
% Stevinweg 1 / PO-box 5048,
% 2628 CN Delft,
% E-mail: \href{mailto:E.C.Slob@tudelft.nl}{E.C.Slob@tudelft.nl};
% \footnotemark[2]Instituto Mexicano del Petróleo,
% Eje Central Lázaro Cárdenas Norte 152,
% Col. San Bartolo Atepehuacan C.P. 07730,
% Ciudad de México, México,
% E-mail: \href{mailto:dieter@werthmuller.org}{Dieter@Werthmuller.org};
% \footnotemark[3]Lamont-Doherty Earth Observatory, Columbia University,
% 305C Oceanography,
% 61 Route 9W, PO Box 1000,
% Palisades NY 10964-8000 US,
% E-mail: \href{mailto:KKey@ldeo.columbia.edu}{KKey@ldeo.columbia.edu}.
}

\author{%
% Dieter Werthmüller\footnotemark[1]\footnotemark[2], %
% Kerry Key\footnotemark[3], and %
% Evert C.\ Slob\footnotemark[1]%
}

\footer{}
\lefthead{}%Werthmüller et al.}
\righthead{Digital filter designing tool}

@@

By substituting $r = e^x$ and $l = e^{-y}$ we get
%
\begin{equation}
  e^x F(e^x) = \int^\infty_{-\infty} f(e^{-y})K(e^{x-y})e^{x-y}\mr{d}y\ .
  \label{eq:filtint}
\end{equation}
%
This can be re-written as a convolution integral and be approximated for an
$N$-point filter by
%
\begin{equation}
  F(r) \approx \sum^N_{n=1} \frac{f(b_n/r) h_n}{r}\ ,
  \label{eq:filtapprox}
\end{equation}
%
where $h$ is the digital linear filter, and the logarithmically spaced filter
abscissae is a function of the spacing $\Delta$ and the shift $\delta$,
%
\begin{equation}
  b_n = \exp\left\{\Delta(-\lfloor{(N+1)/2\rfloor}+n) + \delta\right\} \ .
  \label{eq:base}
\end{equation}
%
\new{Note to reviewers: In equation~\ref{eq:base} $N/2$ changed to
$\lfloor{(N+1)/2\rfloor}$.} From equation~\ref{eq:filtapprox} it can be
seen that the filter method requires $N$ evaluations at each $r$. For example,
to calculate the frequency domain result for 100 offsets with a 201\,pt filter
requires 20,100 evaluations in the wavenumber domain. This is why the DLF often
uses interpolation to minimize the required evaluations, either for $F(r)$ in
what is referred to as \emph{lagged convolution DLF}, or for $f(l)$, which we
here call \emph{splined DLF}.

In the direct matrix inversion method for solving for the digital filter
coefficients, equation~\ref{eq:filtapprox} is cast as a linear system\old{ as
follows.}\new{, where} $\Delta$ and $\delta$ are preassigned scalar values and
$r$ is a range of $M$ preassigned values $r_m$. The filter base coefficients
$b_n$ are computed using equation~\ref{eq:base} and an array of values for $l$
are computed using $l_{mn} = b_n/r_m$. The linear system's left hand side (LHS)
matrix $\bf{A}$ has dimensions $M \times N$ with coefficients \old{$A_{mn} =
f(l_{mn})/r_n$}\new{$A_{mn} = f(l_{mn})/r_m$} and the right hand side (RHS)
vector $\bf{v}$ has elements \old{$v_n = F(r_n)$}\new{$v_m = F(r_m)$}. The
resulting linear system
%
\begin{equation}
  \bf{A}\bf{h} = \bf{v}
  \label{eq:,linearSystem}
\end{equation}
%
can be solved by direct matrix inversion to obtain the vector of filter
coefficients $\bf{h}$. For a given filter length $N$, there are many subjective
choices that go into the practical implementation of this method, including the
choice of the transform pairs $F(r)$ and $f(l)$, as well as the value for $M$
and the particular range and spacing of the values $r_m$. Once values are
chosen for these variables, an optimal filter can be found by a grid search
over $\Delta$ and $\delta$ for values that produce a high quality filter. The
choice of metric for what constitutes a high quality filter is also subjective
and is further discussed below.

@@

The basic steps are as follows:
\begin{enumerate}
  \item Evaluate RHS of check-function $f_\mr{C}$:\newline
    \old{$d_\mr{R} = f_\mr{C}.\mr{rhs}(r)$}\\
    \new{$d_\mr{R} = f_\mr{C}^\mr{rhs}(r)$}
  \item Loop over each value $\Delta_i$, $\delta_j$ (brute force\new{;
      parameters on the left-hand side are overwritten where $i$, $j$
      are not mentioned}):
    \begin{enumerate}
      \item Calculate filter base:\newline
        \old{$b_n = \exp\left\{\Delta_i(-N/2+n) + \delta_j\right\}$}\\
        \new{$b_n = \exp\left\{\Delta_i(-\lfloor{(N+1)/2\rfloor}+n) +
      \delta_j\right\}$}
      \item Get required RHS and LHS evaluation points:\newline
        $r_\mr{I} = p(b, N)$\newline
        $l_\mr{I} = b/r_\mr{I}$
      \item Invert for filter coefficients:\newline
        \old{$J_{ij} = \mr{solve}(f_\mr{I}.\mr{lhs}(l_\mr{I}),
        f_\mr{I}.\mr{rhs}(r_\mr{I}))$}\\
        \new{$J = \mr{solve}\left[f_\mr{I}^\mr{lhs}(l_\mr{I}),
        f_\mr{I}^\mr{rhs}(r_\mr{I})\right]$}
    \item Calculate numerically RHS of check-function $f_\mr{C}$ with current
      filter \old{$J_{ij}$}\new{$J$}:\newline
      \old{$d_\mr{F} = f_\mr{C}.\mr{lhs}(b/r)\cdot J_{ij}\ /\ r$}\\
      \new{$d_\mr{F} = f_\mr{C}^\mr{lhs}(b/r)\cdot J\ /\ r$}
      \item Store minimum recovered amplitude or maximum $r$ where relative
        error is less than the provided error:\newline
        $\chi_{ij} = g\left[
        \mr{argmin}\left(|(d_\mr{F}-d_\mr{R})/d_\mr{R}| >
        \mr{error}\right) -1\right]\ $, where $g$ is either $d_\mr{R}$
        or $1/r$.
    \end{enumerate}
  \item \old{Return filter}\new{Re-calculate (Steps 2a-2c) filter base and}
    coefficients which yield minimum amplitude or maximum $r$ (a local
    minimization can be run to polish the brute-force result):\newline
    \new{$i_\mr{best}$, $j_\mr{best}$ = $\mr{argmin}(\chi)$}\newline
    \new{recalculate $b$, $J$ with $i_\mr{best}$, $j_\mr{best}$}\newline
    \old{return $J\left[\mr{argmin}(\chi)\right] $}\newline
  \item \new{Return filter base and coefficients and error matrix (for visual
    QC):}\newline
    \new{return $b$, $J$, $\chi$}\newline
\end{enumerate}

@@

The presented, free and open-source code \texttt{fdesign} can be used to design
digital linear filters for the Hankel and Fourier transforms (and more
generally for any linear transform) using either analytical transform pairs or
1D subsurface models together with the EM-modeler \texttt{empymod}. The code is
available from GitHub \old{in the add-on repository}\new{as part} of
\texttt{empymod}\old{, called \texttt{empyscripts},} from version
\old{0.2.0}\new{1.7.0} onwards.

@@

% D.W. would like to thank the electromagnetic research group of Aleksandr
% Mousatov at the IMP for fruitful discussions, and Jürg Hunziker for valuable
% feedback which improved this manuscript considerably.
We thank the reviewers for valuable inputs not only regarding the manuscript,
but also regarding the code and its documentation.
